\mythesistitleNOTSYNPOSIS \tnote[fyi]{Synopsis title is ``Generalization of Consecutive Ones Property of Binary Matrices''}

### 2012-01-25 src/020chap2.tex
* Regular edits. See diff for changes.
* Formatting/layout related changes. stuff i'll forget later.
-------
basically:
- added two side printing
- added headers
- added tag/label markers for draft
- made all chapters/chapter*s start on odd page

technically:
- Added contents to Toc
- increased vglue in ded.tex. adjusted one hspace too. to make them fit in.
- began adding short titles for Toc. (chapter 2)
- added twoside option in LoadClass
- adjusted margins in \RequirePackage{geometry} - reduced left margin. otherwise doc bloats up with a lot of margin whitespace on twoside.
- added headings

%\chapter{\CoP\,-- A Survey of Important Results}
\chapter{Consecutive-ones Property -- A Survey of Important Results}
\label{ch:copsurvey}

@@

\begin{definition}[\em Orthogonal sets \cite{n89, mm96,
    mcc04}\footnote{\cite{mcc04} does not use the term ``mutually
    orthogonal'' but refers to the same idea as ``sets that do not
    overlap''. This terminology is also used in other literature like
    \cite{nsnrs09,wlh02}.} ]
  \label{def:orthosets}%
  Let $\cF$ be a set system with universe $U$ and sets $A, B \in \cF$.
  \begin{enumerate}
  \item $A$ and $B$ are said to have a {\em trivial intersection} if
    $A \cap B$ is one of the following.
    \begin{enumerate}[i.]
    \singlespacing
    \item $\emptyset$
    \item $A$
    \item $B$\footnote{In other words, $A$ and $B$ are either disjoint
        or one is the subset of the other.}
    \end{enumerate}
    % \vspace{-5mm}%

@@

  \item $\overline{\cF}$ represents the smallest super set system of
    $\cF$ that is complete\footnote{\cite[Def.~3.2]{mcc04} calls
      this the {\em weak closure} of $\cF$ denoted by $\cW(\cF)$. \label{mcc3}}
  \end{enumerate}
  \dstop
\end{definition}

@@

@@

Proposition~\ref{pr:validcop} is owing to the following theorem by
which \cite{mm96} describes a way to decompose the problem of finding
all \COP orders of $\cF$ into two subproblems using sets in
$\overline{\cF} \cap \cF^{\bot}$.

@@

\def \ff {\overline{\cF} \cap \cF^{\bot}}

The idea behind Theorem~\ref{th:mmdecomp} is as follows. A permutation $\alpha$
of $U$ is a composition of two permutations with respect to $H$ - i) a
permutation $\gamma$ of $H$ and (ii) a permutation $\beta$ of $U/H$.

For two mutually orthogonal sets $A, B$ such that $A \nsubseteq B$,
$A/B$ is defined as the set obtained by removing all elements of $B$
from $A$ and adding a representative element for $B$ in $A$. Being orthogonal,
results in only the following three possibilities.
\begin{enumerate}
\item $A$ and $ B$ are disjoint: $A/B  = A$
\item $A$ is a subset of $B$: $A/B$ is not defined
\item $B$ is a subset of $A$: $A/B = A \setminus B \cup \{b\}$, where $b$
is a new element not in $U$ added to represent $B$.
\end{enumerate}

@@

The following corollary states how \PQRtree elegantly fits into this
whole theory and help in computing all \COP orders of $\cF$.

\begin{corollary}[{\cite[Cor.~8]{mm96}}]
  Let $\cF$ be a set system with universe $U$ and $H$ is a non-empty
  orthogonal set $H \in
  \ff$. If there is a \PQRtree $T_1$ that encodes all permutations in
  $\valid{U/H, \cF/H}$ and also a \PQRtree $T_2'$ that encodes all
  permutations $\valid{H,\cF \cap \power{H}}$, then a \PQRtree $T$ for
  $\cF$ can be obtained by replacing the leaf $h$ in $T_1$ by $T_2$.
\end{corollary}

@@

Thus we have a recursive algorithm that can compute the \PQRtree for
$\cF$ provided we find an element from $\ff$ in each iteration. The
non-empty sets in $\ff$ are called {\em node sets} since they form the
nodes in the \PQRtree. They are calculated as follows.  One is by
computing the overlap components of $\cF$. The overlap components is
the partition that results from the {overlap} equivalence relation
which is nothing but non-orthogonal equivalence relation\footnote{It
  is easy to verify that this relation is indeed an equivalence
  relation.}. Overlap components are linearly
computable\cite{mm95,wlh92}. Once these elements are factored
out, the rest of the node sets are obtained by identifying {\em twin}
elements\footnote{\cite[Sec.~3]{kklv10} calls this {indistinguishable}
  elements. The equivalence class is called a {\em slot}.}. Two
elements $a, b \in U$ are twins if their membership in every set of
$\cF$ is in tandem with each other, \ie $\{a,b\} \bot \cF$. This is
clearly an equivalence relation and their equivalence classes is known
to be computable in linear time\cite{wlh01,mm96-ref11} and even in
logspace\cite{kklv10}.

The recursion end condition is when one cannot find any more sets from
$\ff$ that are non-trivial. This is when $\ff = \cT(U)$. This is the
point where the parents of the leaves of the \PQRtree are created. The
following theorem helps the algorithm decide whether a $P$, $Q$ or
$R$ node must be created.

\begin{theoremsansproof}[{\cite[Th.~9]{mm96},\cite[Th.~2.1]{mcc04}}]
\label{th:pqrfinalmm}
If $\cF$ is a set system with universe $U$ and $|U| \ge 3$ then one of
the following statements hold.
\begin{enumerate}
\item \label{th::it1} $\overline{\cF} = \cT(U)$ 
\item \label{th::it2}$\overline{\cF} = consec(\alpha)$\tnote{what is consec?} for
  some permutation $\alpha$ on $U$ 
\item \label{th::it3}$\overline{\cF} = \power{U}$ 
\end{enumerate}
\end{theoremsansproof}

In case (\ref{th::it1}) all elements in $U$ are made children of a
$P$ node. In case (\ref{th::it2}) all elements in $U$ are made
children of a $Q$ node in the order given by $\alpha$. Finally case
(\ref{th::it3}) is the one when no permutation of $U$ gives \COP. All
elements of $U$ are in this case made children in an $R$ node.

Theorem~\ref{th:pqrfinalmm} and the theory leading to it is very
similar to \cite[Th.~2.1, 3.5. Also Th.~3.2, 3.3, 3.4]{mcc04} which categorizes
the nodes in the above the three cases as {\em prime}, {\em linear}
and {\em degenerate} respectively. Their generalized \PQtree is
created in similar ways as \PQRtree above. This tree is in essense the
Hasse diagram of what they call {\em strong elements} of
$\overline{\cF}$. Strong elements of a set family are elements that do
not overlap with any other elements in the family, \ie it is
orthogonal to all other sets\cite[Def.~3.3]{mcc04}.

@@

As mentioned before, we observe the three data strutures of \PQRtree,
\gPQtree and generalized \PQtree to be equivalentand the theory of all
these data structures is summarised in Table~\ref{tab:pqrcomparison}.

We will now see one more generalization of \PQtree before seeing other
approaches to solving \COP testing including ICPIA which is a set
theoretic characterization of \COP.

@@

@@

\begin{definition}[Intersection Cardinality Preserving Interval
  Assignment (ICPIA)]
  \label{def:icpia}
  Let $\cF = \{A_i \mid A_i \subseteq U, i \in [m]\}$ be a set system
  from universe $U$ and the set of ordered pairs $\Pi = \{(A_i, B_i)
  \mid B_i  \text{ is an interval from } [n], i \in [m]\}$ be an {\em interval
    assigment} of $\cF$, then it is called an {\em ICPIA} if it has
  the following properties.
  \begin{enumerate}[i.]
  \item $|A_i| = |B_i|$ for all $i \in [m]$
  \item $|A_i \cap A_j| = |B_i \cap B_j|$ for all $i, j \in [m]$
  \end{enumerate}
\end{definition}

@@

The necessity of ICPIA for a set system to have \COP given in
Theorem~\ref{th:inpia1} is fairly obvious. It turns out that it is
sufficient too.  This is demostrated by two algorithms which work as
follows. The first one iterates over the assignment
$\Pi$ %set system $\cF$
and changes it in each iteration till the altered
assignment %set system
has no more overlapping intervals\footnote{Intervals are the second
  element in each ordered pair of $\Pi$. Def.~\ref{def:icpia}} %sets
\cite[Alg.~1]{nsnrs09}. This is acheived by replacing two overlapping
assignment pairs $(A_i, B_i)$ and $(A_j, B_j)$ with the partitions of
$(A_i \cup A_j,B_i \cup B_j)$ induced by their overlaps -- $(A_i
\setminus A_j, B_i \setminus B_j)$, $(A_i \cap A_j, B_i \cap B_j)$ and
$(A_j \setminus A_i, B_j \setminus B_i)$.

@@

What we saw so far in this section is checking for the feasibility of
a given interval assignment. ICPIA can also be used to find an
interval assignment when only the set system $\cF$ is given. Part of
the approach is similar to the overlap component idea in \cite{fg65,
  wlh02}. Once the overlap components are factored out, they can be
independently assigned a subinterval from to which the sets in the
component need to be mapped. This is simple because the overlap
components are disjoint from each other. Now the subproblem is finding
\COP order for each overlap component. In each component, the set
whose intersections with the rest of the sets in that component forms
a single inclusion chain is chosen first and assigned the
leftmost\footnote{By convention. Rightmost can also be chosen
  analogously to get a different \COP order.} interval. The next
candidate set is chosen from this set's overlapping sets. The interval
assigned is calculated by simple intersection cardinality
means\footnote{Assume that the next set overlaps on one of the
  ``ends'' of the current set and calculate the interval by shifting
  left or right of the current interval and adjusting its size to
  match the next set's cardinality.}. Once the next set is assigned an
interval, the current assigned sets are all checked for ICPIA. This is
a backtracking algorithm -- if at any point ICPIA fails, another
overlapping set is chosen. If all overlapping set attempts fail ICPIA,
a different ``first set'' is chosen. This is a remarkably simple
algorithm due to its obvious simplicity in implementation.

@@

@@

\temptext{TEXT INTRODUCING SECTION}

@@

\cite{mcc04} describes another characterization for matrices without
\COP which is efficiently computable. This is done by way of an {\em
  incompatibility graph} of a set system\footnote{Since \COP in
  matrices and set systems are equivalent problems as seen in
  Section~\ref{sec:surveycertalgo} we will use the term set system
  here.}. The construction of this graph is based on the following
observation.

\begin{observation} 
  Consider elements $a, b, c \in U$ and a set system $\cF$ with
  universe $U$. If there is at least one set $S \in \cF$ such that $a,
  c \in \cF$ but $b \notin S$, then it is impossible to have a \COP
  order that will place $b$ in between $a$ and $c$. \tnote{also cite
    the inbetweenness problem}
\end{observation}

>>>>

%\chapter{\CoP\,-- A Survey of Important Results}
\chapter[Consecutive-ones Property -- A Survey]{Consecutive-ones Property -- A Survey of Important Results}
\label{ch:copsurvey}

@@

\begin{definition}[\em Orthogonal sets \cite{n89, mm96,
    mcc04}\footnote{\cite{mcc04} does not use the term ``mutually
    orthogonal'' but refers to the same idea as ``sets that do not
    overlap''. This terminology is also used in other literature like
    \cite{nsnrs09,wlh02}.} etc.]
  \label{def:orthosets}%
  Let $\cF$ be a set system with universe $U$ and sets $A, B \in \cF$.
  \begin{enumerate}
  \item $A$ and $B$ are said to have a {\em trivial intersection} if
    $A \cap B$ is one of the following.
    \begin{enumerate}[i.]
    \singlespacing
    \item $\emptyset$
    \item $A$
    \item $B$\footnote{In other words, $A$ and $B$ are either disjoint
        or one is the subset of the other.}
    \end{enumerate}
    % \vspace{-5mm}%

@@

  \item $\overline{\cF}$ represents the smallest super set system of
    $\cF$ that is complete\footnote{\cite[Def.~3.2]{mcc04} calls
      this the {\em weak closure} of $\cF$ denoted by
      $\cW(\cF)$. \label{mcc3}}

  \item \label{def::slashop} The binary operator $/$ is defined as
    follows. Let $A, B$ be two sets from set sysetm $\cF$.
    \begin{align*}
      A/B &= A \setminus B \cup \{b\}\\
      \cF/B &= \{ S/B \mid S \in \cF, S \nsubseteq B \}
    \end{align*}
    $b$ is a new element added as a representative for $B$.
 \end{enumerate}
  \dstop
\end{definition}

@@

\temptext{narrative is abrupt from here. add some explanation for the
  theorem and proposition below?}

@@

Proposition~\ref{pr:validcop} is owing to the following theorem by
which \cite{mm96} describes a way to decompose the problem of finding
all \COP orders of $\cF$ into two subproblems using sets in
$\overline{\cF} \cap \cF^{\bot}$. The operator $/$ used is defined in
Definition~\ref{def:orthosets}~(\ref{def::slashop}).

@@

\def \ff {\overline{\cF} \cap \cF^{\bot}}
The idea behind Theorem~\ref{th:mmdecomp} is as follows. A permutation $\alpha$
of $U$ is a composition of two permutations with respect to $H$ - (i) a
permutation $\gamma$ of $H$ and (ii) a permutation $\beta$ of $U
\setminus H$.

For two mutually orthogonal sets $A, B$ such that $A \nsubseteq B$,
$A/B$ is defined as the set obtained by removing all elements of $B$
from $A$ and adding a representative element for $B$ in $A$. Being orthogonal,
$A/B$ results in only the following three possibilities.
\begin{enumerate}[i. ]
\item $A$ and $ B$ are disjoint: $A/B  = A$
\item $A$ is a subset of $B$: $A/B$ is not defined
\item $B$ is a subset of $A$: $A/B = A \setminus B \cup \{b\}$, where $b$
is a new element not in $U$ added to represent $B$.
\end{enumerate}

@@

The following corollary, Corollary~\ref{cor:pqrrecursive} states how
\PQRtree elegantly fits into this whole theory and help in computing
all \COP orders of $\cF$.

\begin{corollary}[{\cite[Cor.~8]{mm96}}]
  \label{cor:pqrrecursive}
  Let $\cF$ be a set system with universe $U$ and $H$ is a non-empty
  orthogonal set $H \in \ff$. If there is a \PQRtree $T_1$ that
  encodes all permutations in $\valid{U/H, \cF/H}$ and also a \PQRtree
  $T_2$ that encodes all permutations $\valid{H,\cF \cap \power{H}}$,
  then a \PQRtree $T$ for $\cF$ can be obtained by replacing the leaf
  $h$ in $T_1$ by $T_2$, where $h$ is the representative element for $H$.
\end{corollary}

@@

Thus we have a recursive algorithm that can compute the \PQRtree for
$\cF$ provided we find an element from $\ff$ in each iteration. The
non-empty sets in $\ff$ are called {\em node sets} since they form the
nodes in the \PQRtree, as alluded to by
Corollary~\ref{cor:pqrrecursive}. They are calculated as follows.
First method is by computing the overlap components of $\cF$. The
overlap components is the partition that results from the {\em
  overlap} equivalence relation which is nothing but {\em
  non-orthogonal} equivalence relation\footnote{It is easy to verify
  that this relation is indeed an equivalence relation.}. Overlap
components are linearly computable\cite{mm95,wlh92}. Once these
elements are factored out, the rest of the node sets are obtained by
the second method which is identifying {\em twin}
elements\footnote{\cite[Sec.~3]{kklv10} calls this {indistinguishable}
  elements. The equivalence class is called a {\em slot}.}. Two
elements $a, b \in U$ are twins if their membership in every set of
$\cF$ is in tandem with each other, \ie $\{a,b\} \bot \cF$. This is
clearly an equivalence relation and their equivalence classes is known
to be computable in linear time\cite{wlh01,mm96-ref11} and even in
logspace\cite{kklv10}.

The recursion end condition is when one cannot find any more sets from
$\ff$ that are non-trivial. This is when $\ff = \cT(U)$. This is the
point where the parents of the leaves of the final \PQRtree are created. The
following theorem helps the algorithm decide whether a $P$, $Q$ or
$R$ node must be created.

\begin{theoremsansproof}[{\cite[Th.~9]{mm96}}] %,\cite[Th.~2.1]{mcc04}
\label{th:pqrfinalmm}
If $\cF$ is a set system with universe $U$ and $|U| \ge 3$ then one of
the following statements hold.
\begin{enumerate}
\item \label{th::it1} $\overline{\cF} = \cT(U)$ 
\item \label{th::it2}$\overline{\cF} = consec(\alpha)$\tnote{what is consec?} for
  some permutation $\alpha$ on $U$ 
\item \label{th::it3}$\overline{\cF} = \power{U}$ 
\end{enumerate}
\end{theoremsansproof}

In case (\ref{th::it1}) of Theorem~\ref{th:pqrfinalmm} all elements in
$U$ are made children of a $P$ node. In case (\ref{th::it2}) all
elements in $U$ are made children of a $Q$ node in the order given by
$\alpha$. Finally, case (\ref{th::it3}) is the situation where no permutation
of $U$ gives \COP. All elements of $U$ are in this case made children
in an $R$ node.

Theorem~\ref{th:pqrfinalmm} and the theory leading to it is very
similar to \cite[esp. Th.~2.1, 3.5. also Th.~3.2, 3.3, 3.4]{mcc04}
which categorizes the nodes in the above the three cases as {\em
  prime}, {\em linear} and {\em degenerate}
respectively. \cite{mcc04}'s generalized \PQtree is created in similar
ways as \PQRtree above. This tree is, in essence, the Hasse diagram of
what they call {\em strong elements} of weak closure,
$\cW(\cF)$\footnote{\ie $\overline{\cF}$}. Strong elements of a set
family are elements that do not overlap with any other elements in the
family, \ie it is orthogonal to all other sets\cite[Def.~3.3]{mcc04}.

@@

As mentioned before, we observe the three data strutures of \PQRtree,
\gPQtree and generalized \PQtree to be equivalent. The terminologies
of the theory of all these data structures is summarised in
Table~\ref{tab:pqrcomparison}.

We will now see, in brief, one more generalization of \PQtree before
seeing another approaches to solving \COP testing called ICPIA which
is a set theoretic characterization of \COP.

@@

\temptext{narrative is abrupt}

@@

\begin{definition}[Intersection Cardinality Preserving Interval
  Assignment (ICPIA)]
  \label{def:icpia}
  Let $\cF = \{A_i \mid A_i \subseteq U, i \in [m]\}$ be a set system
  from universe $U$ and the set of ordered pairs $\Pi = \{(A_i, B_i)
  \mid B_i  \text{ is an interval from } [n], i \in [m]\}$ be an {\em interval
    assignment} of $\cF$, then $\Pi$ is called an {\em ICPIA} if it has
  the following properties.
  \begin{enumerate}[i.]
  \item $|A_i| = |B_i|$ for all $i \in [m]$
  \item $|A_i \cap A_j| = |B_i \cap B_j|$ for all $i, j \in [m]$
  \end{enumerate}
\end{definition}

@@

The necessity condition of Theorem~\ref{th:icpia1} is fairly
obvious. It turns out that it is sufficient too.  This is demostrated
by two algorithms which work as follows. The first one iterates over
the assignment $\Pi$ %set system $\cF$
and changes it in each iteration till the altered
assignment %set system
has no more overlapping intervals\footnote{Intervals are the second
  element in each ordered pair of $\Pi$. Def.~\ref{def:icpia}} %sets
\cite[Alg.~1]{nsnrs09}. This is acheived by replacing two overlapping
assignment pairs $(A_i, B_i)$ and $(A_j, B_j)$ with the partitions of
$(A_i \cup A_j,B_i \cup B_j)$ induced by their overlaps -- $(A_i
\setminus A_j, B_i \setminus B_j)$, $(A_i \cap A_j, B_i \cap B_j)$ and
$(A_j \setminus A_i, B_j \setminus B_i)$.

@@

What we saw so far in this section is checking for the feasibility of
a given interval assignment. ICPIA can also be used to find an
interval assignment when only the set system $\cF$ is given. Part of
the approach is similar to the overlap component idea in \cite{fg65,
  wlh02}. Once the overlap components are factored out, they can be
independently assigned a subinterval to which the sets in the
component need to be mapped. This is simple because the overlap
components are disjoint from each other. Now the subproblem is finding
\COP order for each overlap component. In each component, the set
whose overlaps with the rest of the sets in that component forms
a single inclusion chain\footnote{A set of sets form a {\em single
  inclusion chain} when the Hasse diagram of their subset relation
forms a single chain. In this particular reference, the set of sets is
the set of overlaps.} is chosen first and assigned the
leftmost\footnote{By convention. Rightmost can also be chosen
  analogously to get a different \COP order.} interval. The next
candidate set is chosen from this set's overlapping sets. The interval
assigned is calculated by simple intersection cardinality
means\footnote{Assume that the next set overlaps on one of the
  ``ends'' of the current set and calculate the interval by shifting
  left or right of the current interval and adjusting its size to
  match the next set's cardinality.}. Once the next set is assigned an
interval, the current assigned sets are all checked for ICPIA. This is
a backtracking algorithm -- if at any point ICPIA fails, another
overlapping set is chosen. If all overlapping set attempts fail ICPIA,
a different ``first set'' is chosen. This is a remarkably simple
algorithm due to its obvious simplicity in implementation.

@@

\temptext{ abrupt: Add a section closing para. does this method (above)
  lead to efficient algo? }

@@

In this section we will describe \cite{d08phd}'s approximation
algorithm for finding forbidden submatrix. We will use $\Tuckerset$ to
refer to the set of Tucker's forbidden submatrices shown in
Figure~\ref{fig:forbiddensubmatrices}.

A corollary to one of Tucker's theorems presented in
Theorem~\ref{th:tuckeratfree} is given in
Corollary~\ref{cor:tuckertheorem}. This is central to \cite{d08phd}'s
algorithm.

\begin{corollary}[{\cite[Cor.~3.1]{d08phd}}]
  \label{cor:tuckertheorem}
  A matrix $M$ has \COP \iff its representing bipartite graph
  $G_M$\footnote{Graph whose half adjacency matrix
    (Def.~\ref{def:xxxx}) is $M$.} does not contain an asteroidal
  triple whose three vertices correspond to columns of M.
\end{corollary}

\temptext{
\cite[Ch.~3]{d08phd} describes two main algorithms - one is an
approximation algorithm and the other one is an exact algorithm.

\begin{theoremsansproof}[{\cite[Th.~3.1]{d08phd}}] 
\label{th:forbiddenalg}
Let $M$ be a $(*,\Delta)$-matrix of size $m
\times n$. A forbidden submatrix from $\Tuckerset$ in $M$ that has a minimum
number of rows can be found in $O(\Delta^3m^2n\dot(m+n^2))$
time. Within the same time, one can also find a forbidden submatrix
from $\Tuckerset$ in $M$ that has a minimum number of columns, a minimum number
of rows and columns, or a minimum number of entries.

\end{theoremsansproof}

@@

\cite{mcc04} describes another characterization for matrices without
\COP which is efficiently computable. This is done by way of an {\em
  incompatibility graph} of a set system\footnote{Since \COP in
  matrices and set systems are equivalent problems as seen in
  Section~\ref{sec:surveycertalgo} we will use the term set system
  in this section, in line with \cite{mcc04}'s convention.}. The construction of this graph is based on the following
observation.

\begin{observation} 
  Consider elements $a, b, c \in U$ and a set system $\cF$ with
  universe $U$. If there is at least one set $S \in \cF$ \stt $a, c
  \in S$ but $b \notin S$, then it is impossible to have a \COP order
  that will place $b$ in between $a$ and $c$. \tnote{also cite the
    inbetweenness problem}
\end{observation}
